---
title: "Pathway_Enrichment.Rmd"
author: "Nika Abravan"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# PHASE 4a: RANKING PD-ASSOCIATED GENES ACROSS CELL TYPES
# ====================================================================

# 1. Load the data manipulation toolbelt
library(dplyr)

# 2. Load all four of  exported CSV files

microglia  <- read.csv("~/Desktop/PD_Project_Fresh/DE_Microglia.csv", row.names = 1)
astrocytes <- read.csv("~/Desktop/PD_Project_Fresh/DE_Astrocytes.csv", row.names = 1)
oligo      <- read.csv("~/Desktop/PD_Project_Fresh/DE_Oligodendrocytes.csv", row.names = 1)
neurons    <- read.csv("~/Desktop/PD_Project_Fresh/DE_Dopaminergic_Neurons.csv", row.names = 1)


# require p_val_adj < 0.05 (significant) AND a log2FC > 0.25 (large effect)
filter_and_rank <- function(df) {
  df %>%
    filter(p_val_adj < 0.05 & abs(avg_log2FC) > 0.25) %>%
    arrange(p_val_adj) #it Sorts them so the  strongest genes are at the top
}

# 4. Apply the filter to all four cell types
microglia_sig  <- filter_and_rank(microglia)
astrocytes_sig <- filter_and_rank(astrocytes)
oligo_sig      <- filter_and_rank(oligo)
neurons_sig    <- filter_and_rank(neurons)

# 5. Print the top 5 most critical broken genes in the Microglia to the console
print("Top 5 Most Significant PD Genes in Microglia:")
head(microglia_sig, 5)
```


```{r, fig.width=10, fig.height=7, message=FALSE, warning=FALSE}
# ====================================================================
# PHASE 4b: PATHWAY ENRICHMENT (MICROGLIA)
# ====================================================================

# 1. Loading  biological analysis libraries 
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)

# 2. Extract the gene names from filtered Microglia list
microglia_genes <- microglia_sig$Gene_Symbol

# 3. Translate the Gene Symbols into Entrez IDs 

microglia_entrez <- mapIds(org.Hs.eg.db, 
                           keys = microglia_genes, 
                           column = "ENTREZID", 
                           keytype = "SYMBOL", 
                           multiVals = "first")

# cleaning any genes that didn't translate perfectly
microglia_entrez <- na.omit(microglia_entrez)

# 4. Run the Gene Ontology (GO) Biological Process Enrichment
# This asks the database: "What systems do these broken genes belong to?"
microglia_go <- enrichGO(gene          = microglia_entrez,
                         OrgDb         = org.Hs.eg.db,
                         ont           = "BP", # BP = Biological Process
                         pAdjustMethod = "BH",
                         pvalueCutoff  = 0.05,
                         readable      = TRUE)

# 5. plot 
# This plots top 15 malfunctioning biological systems
dotplot(microglia_go, showCategory = 15) + 
  ggtitle("Microglia: Top PD-Associated Biological Pathways") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.y = element_text(size = 11)) # make it cleaner graph
```


```{r, message=FALSE, warning=FALSE}
# ====================================================================
# PHASE 4d: GSEA (dopaminergic neurons)

library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)

# 1. Load the full dataset
neurons_full <- read.csv("~/Desktop/PD_Project_Fresh/DE_Dopaminergic_Neurons.csv", row.names = 1)

# 2. Translate to Entrez IDs
neurons_full$Entrez <- mapIds(org.Hs.eg.db, 
                              keys = neurons_full$Gene_Symbol, 
                              column = "ENTREZID", 
                              keytype = "SYMBOL", 
                              multiVals = "first")

# Remove any genes that didn't translate
neurons_full <- neurons_full[!is.na(neurons_full$Entrez), ]

# 3. THE FIX: Collapse duplicate barcodes by averaging their log2FC
neurons_clean <- neurons_full %>%
  group_by(Entrez) %>%
  summarize(avg_log2FC = mean(avg_log2FC, na.rm = TRUE)) %>%
  ungroup()

# 4. Build the strictly unique Ranked List
ranked_list <- neurons_clean$avg_log2FC
names(ranked_list) <- as.character(neurons_clean$Entrez)
ranked_list <- sort(ranked_list, decreasing = TRUE)

# 5. Run the GSEA Math 
# We set seed so the random permutations are exactly reproducible
set.seed(1234) 
neurons_gsea <- gseGO(geneList     = ranked_list,
                      OrgDb        = org.Hs.eg.db,
                      ont          = "BP", 
                      minGSSize    = 10,
                      maxGSSize    = 500,
                      pvalueCutoff = 0.05,
                      verbose      = FALSE)

# 6. Draw the GSEA Plot for Noah
dotplot(neurons_gsea, showCategory = 10, split=".sign") + facet_grid(.~.sign) +
  ggtitle("Dopaminergic Neurons: GSEA Pathway Analysis") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```



```{r, fig.width=10, fig.height=7, message=FALSE, warning=FALSE}
# ====================================================================
# PHASE 4e: PATHWAY ENRICHMENT (ASTROCYTES)

# 1. Load the biological analysis toolbelt
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(dplyr)

# 2. Load the Astrocytes CSV
astrocytes_de <- read.csv("~/Desktop/PD_Project_Fresh/DE_Astrocytes.csv", row.names = 1)

# 3. Filter for ONLY the significantly altered genes
astrocytes_sig <- astrocytes_de %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) > 0.25)

# 4. Extract just the gene names
astrocytes_genes <- astrocytes_sig$Gene_Symbol

# 5. Translate the Gene Symbols into Entrez IDs 
astrocytes_entrez <- mapIds(org.Hs.eg.db, 
                            keys = astrocytes_genes, 
                            column = "ENTREZID", 
                            keytype = "SYMBOL", 
                            multiVals = "first")

# Clean up genes that didn't translate perfectly
astrocytes_entrez <- na.omit(astrocytes_entrez)

# 6. Run the Gene Ontology (GO) Biological Process Enrichment
astrocytes_go <- enrichGO(gene          = astrocytes_entrez,
                          OrgDb         = org.Hs.eg.db,
                          ont           = "BP", # BP = Biological Process
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.05,
                          readable      = TRUE)

# 7. Plot 
dotplot(astrocytes_go, showCategory = 15) + 
  ggtitle("Astrocytes: Top PD-Associated Biological Pathways") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```


```{r, message=FALSE, warning=FALSE}
# ====================================================================
# PHASE 4f: PATHWAY ENRICHMENT (OLIGODENDROCYTES)
# ====================================================================

# 1. Load the toolbelt
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(dplyr)

# 2. Load the Oligodendrocytes CSV 
oligo_de <- read.csv("~/Desktop/PD_Project_Fresh/DE_Oligodendrocytes.csv", row.names = 1)

# 3. Filter for ONLY the significantly altered genes
oligo_sig <- oligo_de %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) > 0.25)

# 4. Extract just the gene names
oligo_genes <- oligo_sig$Gene_Symbol

# 5. Translate the Gene Symbols into Entrez IDs 
oligo_entrez <- mapIds(org.Hs.eg.db, 
                       keys = oligo_genes, 
                       column = "ENTREZID", 
                       keytype = "SYMBOL", 
                       multiVals = "first")

oligo_entrez <- na.omit(oligo_entrez)

# 6. Run the Enrichment
oligo_go <- enrichGO(gene          = oligo_entrez,
                     OrgDb         = org.Hs.eg.db,
             ont           = "BP", 
             pAdjustMethod = "BH",
             pvalueCutoff  = 0.05,
             readable      = TRUE)

# 7. plot 
dotplot(oligo_go, showCategory = 15) + 
  ggtitle("Oligodendrocytes: Top PD-Associated Biological Pathways") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```


```{r, message=FALSE, warning=FALSE}

# PHASE 5: BIOLOGICAL SYNTHESIS TABLE (UPGRADED)

# PHASE 5: BIOLOGICAL SYNTHESIS TABLE
# ====================================================================

# 1. Load the table rendering library
library(knitr)

# 2. Build the data table row by row
pd_summary <- data.frame(
  Cell_Type = c("Microglia", 
                "Dopaminergic Neurons", 
                "Astrocytes", 
                "Oligodendrocytes"),
  
  Core_Biological_Failure = c("Lysosomal Overload / Garbage Disposal", 
                              "Synaptic & Axonal Breakdown", 
                              "Lipid / Cholesterol Supply Chain", 
                              "Myelin Maintenance & Production"),
  
  Top_Implicated_Pathways = c("Response to unfolded protein, Lysosome localization", 
                              "Modulation of chemical synaptic transmission", 
                              "Glycerolipid & cholesterol metabolic process", 
                              "Phospholipid metabolic process, Axonogenesis")
)

# 3. Print the table beautifully
kable(pd_summary, 
      col.names = c("Cell Type", "Biological Failure", "Top Implicated Pathways"),
      caption = "Summary of PD-Associated Biological Failures by Cell Type",
      align = "l")

```

### Phase 4 Validation: Donor-Level Pseudobulk Results

To ensure the single-cell pathway signatures are robust across individual patients, pseudobulk validation was performed using DESeq2. The raw counts were aggregated by Patient ID and Cell Type. 

Below are the top significant genes (padj < 0.05) driving the disease signatures in Astrocytes and OPCs.

```{r, message=FALSE, warning=FALSE}
# Load the pre-calculated DESeq2 results
astro_results <- read.csv("~/Desktop/PD_Project_Fresh/Astrocytes_Pseudobulk_Results.csv")
opc_results <- read.csv("~/Desktop/PD_Project_Fresh/OPCs_Pseudobulk_Results.csv")

# Print clean tables for the HTML report
knitr::kable(head(astro_results, 10), 
             caption = "Top 10 Significant Genes in Astrocytes (PD vs Control)")

knitr::kable(head(opc_results, 10), 
             caption = "Top 10 Significant Genes in OPCs (PD vs Control)")
```

### Phase 5: Pathway Enrichment (Gene Ontology)

To understand the broader biological impact of the significant genes identified in our pseudobulk validation, we performed Gene Ontology (GO) enrichment analysis. The bubble plot below highlights the top biological processes disrupted in Parkinson's disease Astrocytes, specifically pointing to protein misfolding and metabolic stress.

```{r, fig.width=8, fig.height=7, message=FALSE, warning=FALSE}
# Load the pathway libraries
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)

# Load the Astrocytes results we saved earlier
astro_results <- read.csv("Astrocytes_Pseudobulk_Results.csv")

# Filter for genes using the raw p-value to cast a wider net
sig_astro_broad <- subset(astro_results, pvalue < 0.05 & !is.na(Gene_Symbol))
significant_genes_broad <- sig_astro_broad$Gene_Symbol

# Run the Gene Ontology (GO) Enrichment Math
ego_astro_broad <- enrichGO(gene          = significant_genes_broad,
                            OrgDb         = org.Hs.eg.db,
                            keyType       = "SYMBOL",
                            ont           = "BP",  
                            pAdjustMethod = "BH",
                            pvalueCutoff  = 0.05,
                            qvalueCutoff  = 0.05)

# Generate the Bubble Plot
dotplot(ego_astro_broad, showCategory = 10, title = "Top Enriched Pathways in Astrocytes (PD vs Control)")
```
